\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-24 paper/algorithm.tex
paper: added similar algo breakout for Time as for Loss

For each sub-plan, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
that plan. We use this estimate to compute $\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     \{1\} \cup \textsc{L}(q') & q = \delta_C(q') \\
    \left\{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}}\right\} \cup \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_1') \cup \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
    \emptyset & q = t\\
  \end{cases}\\
  \textsc{Loss}(q) &= \frac{\Sigma \textsc{L}(q)}{|\textsc{L}(q)|}
\end{align*}

\todo{Similar algorithm for computing Time?}

@@

To compute $\text{Time}(Q)$, we retain a simple set of heuristics used by a standard 
database. Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

@@

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (Lines~\ref{lst:line:attr-start}-\ref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (Line~\ref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (Line~\ref{lst:line:join})\todo{Citation or explanation for join optimization algorithm}. This set is now extended for any necessary grouping and aggregation
operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

@@

\begin{algorithm}
    \input{algorithms/cart.tex}
\end{algorithm}

@@

By restricting imputations to two classes: those that impute only attributes used in the operator and those that impute all attributes needed downstream in the query,
we limit the factor of plans at each operator to three: \textit{Drop}, \textit{Impute} over attributes used in operator, and \textit{Impute} over all attributes needed.
Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set. So in many
cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
then we could explore $\mathcal{O}(3^{\# of operators}) * plans$\todo{this is unclear. what is plans?}.

>>>>

For each sub-plan, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
that plan. We use this estimate to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     \{1\} \cup \textsc{L}(q') & q = \delta_C(q') \\
    \left\{\frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}}\right\} \cup \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_1') \cup \textsc{L}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
    \emptyset\ multiset & q = t\\
  \end{cases}\\
  \textsc{Loss}(q) &= \frac{\Sigma \textsc{L}(q)}{|\textsc{L}(q)|}
\end{align*}

@@

To compute $\text{Time}(Q)$, an additive measure of runtime cost, we retain a simple set of heuristics used by a standard 
database.

%\begin{align*}
%  \textsc{T}(q) &= \begin{cases}
%     \{ f_1(pg_{fetch}, io_{cost}) \} \cup \textsc{T}(q') & q = \delta_C(q') \\
%      \{ f_2(\textsc{Attr}(q'), C, \textsc{Card}(q'), iters) \} \cup \textsc{T}(q')  & q = \mu_C(q') \\  
%     \{ f_3(\textsc{T}(q_1'), \textsc{T}(q_2')) \} & q = q_1' \bowtie_\psi q_2' \\
%    \{ \hat{t} \} \cup \textsc{T}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
%    \{ f_4(\textsc{Card}(t), pg_{size}, io_{cost}) \} & q = t\\
%  \end{cases}
%  \intertext{ \textit{for some set of functions  $f_{1,2,3,4}$, constant $\hat{t}$, IO Cost/Page $io_{cost}$, page fetches $pg_{fetch}$, page size $pg_{size}$, iterations $iters$} } 
%  \textsc{Time}(q) &= \Sigma \textsc{T}(q)
%\end{align*}

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

@@

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (Lines~\ref{lst:line:attr-start}-\ref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (Line~\ref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (Line~\ref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which maintains the Pareto frontier of possible plans. The output of the join optimizer is the set of plans which satisfy all join predicates and lie along the frontier.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
epochs) or a fixed number of epochs is reached.

@@

%\begin{algorithm}
%    \input{algorithms/cart.tex}
%\end{algorithm}

@@

By restricting imputations to two classes: those that impute only attributes used in the operator and those that impute all attributes needed downstream in the query,
we limit the factor of plans at each operator to three: \textit{Drop}, \textit{Impute} over attributes used in operator, and \textit{Impute} over all attributes needed.
Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set. So in many
cases we can drop some of these intermediate plans generated at each operator. However, if the dirty sets tracked are distinct through the entire planning phase,
then we could explore exponentially more plans than System R.
